---
title: "Forecasting U.S. Gasoline Prices under Geopolitical Shocks"
author: "Team ShohanXuKimKim"
date: 04-15-2026
format: 
  html:
    toc: true
    theme: cosmo
execute:
  warning: false
  message: false
editor: visual
---

## 1. Project Overview

We want to analyze how major geopolitical shocks affect U.S. gasoline prices and compare different time series models to see which performs best under these conditions.

-   **Russia–Ukraine War** (Feb 2022)
-   **Israel–Iran Conflict** (June 2025)
-   **US Intervention in Venezuela** (Jan 2026)

## 2. Data Loading & Cleaning

Here we load the raw datasets (Gas prices, WTI prices, and Gas stocks) from the local folder and handle missing values and outliers.

```{r}
# Load required packages
library(tidyverse)
library(lubridate)
library(timetk)

# Read Data from Local 'Raw_Data' Folder
gas_price_df <- read_csv("Raw_Data/Weekly U.S. Gas Price (EIA).csv", skip = 4) %>%
  rename(date = 1, gas_price = 2) %>% mutate(date = mdy(date))

wti_price_df <- read_csv("Raw_Data/Weekly WTI Price (EIA).csv", skip = 4) %>%
  rename(date = 1, wti_price = 2) %>% mutate(date = mdy(date))

gas_stocks_df <- read_csv("Raw_Data/Weekly U.S. Gas Stocks (EIA).csv", skip = 4) %>%
  rename(date = 1, gas_stocks = 2) %>% mutate(date = mdy(date))

# Merge and Clean Data
clean_data <- gas_price_df %>%
  left_join(wti_price_df, by = "date") %>%
  left_join(gas_stocks_df, by = "date") %>%
  arrange(date) %>%
  filter(date >= as.Date("2018-01-01")) %>%
  mutate(
    gas_price = ts_clean_vec(ts_impute_vec(gas_price, period = 1), period = 1),
    wti_price = ts_clean_vec(ts_impute_vec(wti_price, period = 1), period = 1),
    gas_stocks = ts_clean_vec(ts_impute_vec(gas_stocks, period = 1), period = 1)
  )

# Preview the cleaned data
head(clean_data)
```


## 3. Exploratory Data Analysis (EDA)

And we visualize the cleaned gasoline prices.

```{r}
# Plot the target variable
clean_data %>%
  plot_time_series(date, gas_price, 
                   .interactive = TRUE, 
                   .title = "Cleaned U.S. Gasoline Prices")
```

